---
title: "HW-2"
author: "Priyankkumar Chandrakant Patel"
date: "2026-02-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Part A — Data Story Dataset + Visualization (Food Deserts)

## Problem 1 — Import + processing (same as Data Story)

```{r libraries}
library(tidyverse)  # readr, dplyr, ggplot2, tibble, etc.
library(scales)     # comma() formatting
library(patchwork)  # combine plots side-by-side
```

```{r import-data}
# Primary source: USDA ERS Food Access Research Atlas (2019 release)
# Note: the CSV uses the literal string "NULL" for missing values
df_raw <- read_csv("Food Access Research Atlas.csv",
                   na = "NULL",
                   show_col_types = FALSE)

head(df_raw, 10)
```

### Dataset source (primary)

- **Dataset:** USDA Economic Research Service (ERS), *Food Access Research Atlas* (2019 release)
- **Source page:** https://www.ers.usda.gov/data-products/food-access-research-atlas/
- **Accessed via:** Data.gov listing of the same USDA ERS dataset (primary U.S. Government open data portal)

### Why I chose this dataset (aim)

I chose the USDA Food Access Research Atlas because it measures a very real, everyday problem: **how far people live from a supermarket and whether they have the resources (income + transportation) to reach one**.  
My aim was to answer one clear question:

> **How large is the “food desert” problem in the U.S. — in both number of census tracts and number of people affected?**

This makes the visualization an explanatory “scale” story: it quantifies the problem before exploring deeper patterns.

### Variables of interest (used for the visualization)

For the main “scale of the problem” figure (Problem 2), I used:

- **`LILATracts_1And10`**: USDA indicator for “low-income + low-access” census tracts (food desert definition used in this story)
- **`Pop2010`**: tract population (2010 Census)

I also created a derived label:

- **`food_desert`**: `"Food Desert"` if `LILATracts_1And10 == 1`, else `"Not Food Desert"`

### Data processing steps

```{r processing}
# 1) Create a readable label for whether a tract is a food desert (per USDA LILA flag)
df <- df_raw %>%
  mutate(
    food_desert = factor(LILATracts_1And10,
                         levels = c(0, 1),
                         labels = c("Not Food Desert", "Food Desert"))
  )

# 2) Compute totals used in the visualization
total_pop <- sum(df$Pop2010, na.rm = TRUE)

# Population in food deserts = sum of tract population where the LILA flag is 1
fd_pop <- df %>%
  filter(LILATracts_1And10 == 1) %>%
  summarise(pop = sum(Pop2010, na.rm = TRUE)) %>%
  pull(pop)

# Number of food desert tracts
fd_tracts <- sum(df$LILATracts_1And10 == 1, na.rm = TRUE)

total_pop
fd_pop
fd_tracts
```

---

## Problem 2 — Reproduce the visualization + interpretation

```{r visualization, fig.width=12, fig.height=5}
# Panel A: Count of tracts by food desert status
p1 <- df %>%
  count(food_desert) %>%
  ggplot(aes(x = food_desert, y = n, fill = food_desert)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = format(n, big.mark = ",")),
            vjust = -0.5, fontface = "bold", size = 4.5) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title    = "1 in 8 Census Tracts Is a Food Desert",
    subtitle = paste0(format(fd_tracts, big.mark = ","), " of ",
                      format(nrow(df), big.mark = ","), " tracts meet the USDA LILA definition"),
    x = NULL,
    y = "Number of Census Tracts"
  ) +
  theme(legend.position = "none")

# Panel B: Population donut (food desert residents vs rest of US)
pop_df <- tibble(
  category = c("Rest of US", "Food Desert Residents"),
  pop      = c(total_pop - fd_pop, fd_pop)
)

p2 <- ggplot(pop_df, aes(x = 2, y = pop, fill = category)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void(base_size = 13) +
  labs(
    title    = "~39 Million People Live in Food Deserts",
    subtitle = paste0(round(fd_pop / total_pop * 100, 1), "% of the US population"),
    fill = NULL
  ) +
  theme(legend.position = "bottom")

p1 + p2 + plot_annotation(caption = "Source: USDA ERS Food Access Research Atlas (2019)")
```

### Interpretation / key conclusions

- **Scale (tracts):** About **1 in 8 census tracts** are classified as food deserts under the USDA low-income + low-access definition (`LILATracts_1And10`).  
- **Scale (people):** Roughly **39 million people** live in those tracts (based on `Pop2010`).  
- **Main takeaway:** Food deserts are not a “rare edge case” — they affect a population comparable to an entire large country, meaning grocery access is a structural public-health and equity issue rather than an individual choice problem.


#Part B

```{r Loading the tsv dataset and library}
library(tidyverse)




ncaa = read_tsv("26801-0001-Data.tsv", na = c("", "NA", ".", "-9", "-99", "-999"))
```
```{r}
glimpse(ncaa)
```
```{r}

ncaa_selected <- ncaa %>%
  select(SCL_UNITID, SCL_NAME, SPORT_CODE, SPORT_NAME, 
         starts_with("APR_RATE_"))

glimpse(ncaa_selected)
```
```{r}
# Pivot the APR columns into rows
ncaa_long <- ncaa_selected %>%
  pivot_longer(
    cols = starts_with("APR_RATE_"),
    names_to = "year_col",
    values_to = "APR"
  )

# Check the result
head(ncaa_long, 10)
```
```{r}
# Extract just the year number from year_col
ncaa_with_year <- ncaa_long %>%
  mutate(Year = as.integer(str_sub(year_col, 10, 13)))

# Check it worked
head(ncaa_with_year, 10)
```
```{r}
# Remove the intermediate column and rename
ncaa_tidy <- ncaa_with_year %>%
  select(-year_col) %>%
  rename(
    school_id = SCL_UNITID,
    school_name = SCL_NAME,
    sport_code = SPORT_CODE,
    sport_name = SPORT_NAME
  )

# Final check
head(ncaa_tidy, 10)
```
```{r}
ncaa_tidy
```

```{r}
# Boxplot showing APR distribution for each year
ggplot(ncaa_tidy, aes(x = factor(Year), y = APR)) +
  geom_boxplot() +
  labs(
    title = "Distribution of APRs Over Time (2004-2014)",
    x = "Year",
    y = "APR Score"
  ) +
  theme_minimal()
```

```{r}
ggplot(ncaa_tidy, aes(x = factor(Year), y = APR)) +
  geom_violin() +
  theme_minimal()
```

```{r}
ncaa_gendered <- ncaa_tidy %>%
  filter(sport_code >= 1 & sport_code <= 37) %>%
  mutate(gender = ifelse(sport_code <= 18, "Men's", "Women's"))

# Check the result
head(ncaa_gendered, 10)
```
```{r}
ncaa_gendered
```

```{r}
ggplot(ncaa_gendered, aes(x = factor(Year), y = APR, fill = gender)) +
  geom_boxplot() +
  labs(
    title = "Distribution of APRs Over Time by Gender (2004-2014)",
    x = "Year",
    y = "APR Score",
    fill = "Gender"
  ) +
  theme_minimal()
```

```{r}
ggplot(ncaa_gendered, aes(x = factor(Year), y = APR, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~gender) +
  labs(
    title = "Distribution of APRs Over Time by Gender (2004-2014)",
    x = "Year",
    y = "APR Score",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Angle the x-axis labels
```
```{r}
ncaa_gendered %>%
  group_by(Year, gender) %>%
  summarize(mean_APR = mean(APR, na.rm = TRUE)) %>%
  ggplot(aes(x = Year, y = mean_APR, color = gender)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Average APR Over Time by Gender (2004-2014)",
    x = "Year",
    y = "Mean APR Score",
    color = "Gender"
  ) +
  theme_minimal()
```
```{r}
ncaa_sports <- ncaa_gendered %>%
  mutate(sport = str_remove(sport_name, "Men's "),
         sport = str_remove(sport, "Women's "))

ncaa_sports
```

`````{r, fig.width = 12, fig.height = 10}
ggplot(ncaa_sports, aes(x = gender, y = APR, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~sport) +
  labs(
    title = "APR Distribution by Sport and Gender",
    subtitle = "Only sports with both men's and women's teams",
    x = "Gender",
    y = "APR Score",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
`````



```{r, fig.width = 10, fig.height = 14}
ggplot(ncaa_sports, aes(x = gender, y = APR, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~sport, ncol = 3) +
  labs(
    title = "APR Distribution by Sport and Gender",
    subtitle = "Only sports with both men's and women's teams",
    x = "Gender",
    y = "APR Score",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

