---
title: "HW-2"
author: "Priyankkumar Chandrakant Patel"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Part A

## Problem 1

### Dataset Import and Processing

```{r load-libraries-partA}
library(readr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(scales)
library(patchwork)
library(viridis)
```

```{r import-data}
# Import the USDA Food Access Research Atlas dataset
# "NULL" strings in the file represent missing data
data <- read_csv("Food Access Research Atlas.csv",
                 na = "NULL",
                 show_col_types = FALSE)

# Display first 10 rows
head(data, 10)
```

**Dataset:** USDA Economic Research Service — Food Access Research Atlas (2019)

**Source Citation:** U.S. Department of Agriculture, Economic Research Service. *Food Access Research Atlas.* Available at: <https://www.ers.usda.gov/data-products/food-access-research-atlas/>. Downloaded from: <https://catalog.data.gov/dataset> (U.S. Government Open Data Portal).

**Why I chose this dataset:** Food access is one of those problems that hides in plain sight. We hear about hunger in abstract statistics, but rarely see *where* it concentrates, *who* it affects disproportionately, and what surprising factors — like car ownership — amplify the crisis. As someone interested in data-driven insights into public health and social equity, I wanted to explore the geographic, demographic, and socioeconomic patterns behind America's food deserts. The dataset is unusually rich: it covers 72,531 census tracts with variables spanning income, race, vehicle access, SNAP usage, and distance to the nearest supermarket — allowing multi-dimensional exploration of a single social problem.

**Description:** The dataset maps food access indicators across `r format(nrow(data), big.mark = ",")` census tracts in all 50 states, Washington D.C., and U.S. territories. It combines 2010 Census population data with 2015 supermarket location data to identify "food deserts" — census tracts that are both low-income and far from supermarkets.

**Variables of Interest:**

| Variable | Description |
|---|---|
| `LILATracts_1And10` | **Food desert flag** — 1 if the tract is low-income AND the nearest supermarket is >1 mile (urban) or >10 miles (rural) |
| `PovertyRate` | % of tract population below the federal poverty line |
| `MedianFamilyIncome` | Median family income in the tract |
| `Urban` | 1 = urban tract, 0 = rural tract |
| `TractHUNV` | Count of housing units without a vehicle |
| `OHU2010` | Total occupied housing units (denominator for vehicle/SNAP rates) |
| `TractSNAP` | Count of housing units receiving SNAP (food stamp) benefits |
| `Pop2010` | Total tract population from the 2010 Census |
| `TractKids`, `TractSeniors` | Children and senior population counts |
| `TractBlack`, `TractWhite`, `TractHispanic`, `TractAIAN` | Race/ethnicity population counts |

**Data Processing Steps:**

```{r data-processing}
# Create derived columns for analysis
data <- data %>%
  mutate(
    food_desert    = factor(LILATracts_1And10,
                            levels = c(0, 1),
                            labels = c("Not Food Desert", "Food Desert")),
    area_type      = factor(Urban,
                            levels = c(0, 1),
                            labels = c("Rural", "Urban")),
    poverty_bin    = cut(PovertyRate,
                         breaks = c(0, 10, 20, 30, 50, 100),
                         labels = c("0-10%", "10-20%", "20-30%", "30-50%", "50%+"),
                         include.lowest = TRUE),
    # Safe percentage calculations — guard against OHU2010 == 0 and cap at 100
    pct_no_vehicle = if_else(OHU2010 > 0,
                             pmin(TractHUNV / OHU2010 * 100, 100),
                             NA_real_),
    pct_snap       = if_else(OHU2010 > 0,
                             pmin(TractSNAP / OHU2010 * 100, 100),
                             NA_real_),
    pct_kids       = if_else(Pop2010 > 0,
                             TractKids / Pop2010 * 100,
                             NA_real_),
    pct_seniors    = if_else(Pop2010 > 0,
                             TractSeniors / Pop2010 * 100,
                             NA_real_),
    pct_black      = if_else(Pop2010 > 0,
                             TractBlack / Pop2010 * 100,
                             NA_real_),
    pct_white      = if_else(Pop2010 > 0,
                             TractWhite / Pop2010 * 100,
                             NA_real_),
    pct_hispanic   = if_else(Pop2010 > 0,
                             TractHispanic / Pop2010 * 100,
                             NA_real_),
    pct_aian       = if_else(Pop2010 > 0,
                             TractAIAN / Pop2010 * 100,
                             NA_real_)
  )

# Summary statistics
cat("Total tracts:", nrow(data), "\n")
cat("Food desert tracts:", sum(data$LILATracts_1And10 == 1, na.rm = TRUE), "\n")
cat("Population in food deserts:",
    format(sum(data$Pop2010[data$LILATracts_1And10 == 1], na.rm = TRUE), big.mark = ","), "\n")
```

Processing steps performed:

1. Converted binary indicators (`LILATracts_1And10`, `Urban`) to descriptive factor variables (`food_desert`, `area_type`) for clearer visualization labels.
2. Binned `PovertyRate` into five categories for grouped comparisons.
3. Calculated percentage variables (`pct_no_vehicle`, `pct_snap`, `pct_kids`, `pct_seniors`, racial percentages) by dividing count variables by the appropriate denominators (`OHU2010` for housing-based metrics, `Pop2010` for population-based metrics), with safeguards against division by zero and values capped at 100%.

---

## Problem 2

### Visualization: Urban vs. Rural Food Desert Rates and Vehicle Access

The visualization below combines two perspectives into a single figure using `patchwork`: (1) food desert rates by urban vs. rural classification, and (2) the vehicle access gap between food desert and non-food desert tracts. Together, these reveal that food deserts are overwhelmingly an *urban* phenomenon and that vehicle-lessness is a powerful hidden multiplier.

```{r problem2-visualization, fig.width = 10, fig.height = 10}
# --- Panel 1: Food Desert Rate by Urban vs. Rural ---
urban_rural_summary <- data %>%
  group_by(area_type) %>%
  summarise(
    total_tracts = n(),
    fd_tracts    = sum(LILATracts_1And10 == 1, na.rm = TRUE),
    fd_rate      = fd_tracts / total_tracts * 100,
    total_pop    = sum(Pop2010, na.rm = TRUE),
    fd_pop       = sum(Pop2010[LILATracts_1And10 == 1], na.rm = TRUE),
    .groups = "drop"
  )

p1 <- ggplot(urban_rural_summary,
             aes(x = area_type, y = fd_rate, fill = area_type)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(fd_rate, 1), "%\n(",
                                format(fd_tracts, big.mark = ","), " tracts)")),
            vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("Rural" = "#27ae60", "Urban" = "#8e44ad")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2)),
                     labels = label_percent(scale = 1)) +
  labs(title    = "The Urban Surprise: Cities Have Nearly 2x the Food Desert Rate",
       subtitle = "Percentage of census tracts classified as food deserts",
       x = NULL, y = "Food Desert Rate (%)",
       caption  = "Source: USDA Food Access Research Atlas (2019)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))

# --- Panel 2: Vehicle Access Comparison ---
vehicle_summary <- data %>%
  filter(!is.na(pct_no_vehicle)) %>%
  group_by(food_desert) %>%
  summarise(
    mean_no_vehicle = mean(pct_no_vehicle, na.rm = TRUE),
    median_no_vehicle = median(pct_no_vehicle, na.rm = TRUE),
    .groups = "drop"
  )

p2 <- ggplot(data %>% filter(!is.na(pct_no_vehicle)),
             aes(x = food_desert, y = pct_no_vehicle, fill = food_desert)) +
  geom_boxplot(width = 0.5, outlier.alpha = 0.1, outlier.size = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "red") +
  stat_summary(fun = mean, geom = "text",
               aes(label = paste0("Mean: ", round(after_stat(y), 1), "%")),
               vjust = -1.5, size = 3.5, color = "red") +
  scale_fill_manual(values = c("Not Food Desert" = "#3498db",
                                "Food Desert"     = "#e74c3c")) +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title    = "The Car Equation: No Vehicle = 4x Higher Food Desert Risk",
       subtitle = "Distribution of % housing units without a vehicle",
       x = NULL, y = "% Housing Units Without a Vehicle",
       caption  = "Source: USDA Food Access Research Atlas (2019)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 14))

# Combine using patchwork
p1 / p2 +
  plot_annotation(
    title = "America's Food Deserts: An Urban Crisis Amplified by Vehicle Access",
    theme = theme(plot.title = element_text(face = "bold", size = 16))
  )
```

### Interpretation

This combined visualization reveals two key findings from the USDA Food Access Research Atlas:

**Panel 1 — The Urban Surprise:** Contrary to the common stereotype of food deserts as a rural problem, urban census tracts have a food desert rate of approximately 14.4%, nearly double the rural rate of approximately 7.8%. This means that of the roughly 8,500 food desert tracts nationwide, the majority are in cities — where high population density means millions more people are affected.

**Panel 2 — The Car Equation:** Vehicle access is a dramatic differentiator. Food desert tracts show roughly 4 times the mean percentage of housing units without a vehicle compared to non-food-desert tracts. The boxplot shows that the interquartile range for food desert tracts sits much higher, with many tracts where over 20% of households lack a car. In areas with no nearby grocery store, lacking a vehicle transforms inconvenience into genuine food insecurity.

**Key conclusions:**

- Food deserts are primarily an urban phenomenon — debunking the "empty countryside" stereotype.
- Vehicle access acts as a hidden multiplier: even in low-income areas, a car can mean the difference between access and isolation.
- Policy solutions must address both the supply side (opening stores) and the demand side (transportation, delivery infrastructure).
- The concentration of vehicle-less households in food deserts suggests that improving public transit grocery access or subsidizing delivery services could be as impactful as opening new supermarkets.
